---
title: "practical_exercise_2, Methods 3, 2021, autumn semester"
author: "Mille Bryske"
date: "29th of September"
output:
  html_document:
    df_print: paged
---

<style type="text/css">
  body{
  font-size: 14pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment 1: Using mixed effects modelling to model hierarchical data
In this assignment we will be investigating the _politeness_ dataset of Winter and Grawunder (2012) and apply basic methods of multilevel modelling. 

## Dataset
The dataset has been shared on GitHub, so make sure that the csv-file is on your current path. Otherwise you can supply the full path.

```{r}
setwd("/Users/millebryske/Desktop/Uni/3. semester/Methods 3/github_methods_3/week_02")
pacman::p_load(tidyverse)

politeness <- read.csv('politeness.csv') ## read in data
politeness <- na.omit(politeness)
```

# Exercises and objectives
The objectives of the exercises of this assignment are:  
1) Learning to recognize hierarchical structures within datasets and describing them  
2) Creating simple multilevel models and assessing their fitness  
3) Write up a report about the findings of the study  

REMEMBER: In your report, make sure to include code that can reproduce the answers requested in the exercises below  
REMEMBER: This assignment will be part of your final portfolio

## Exercise 1 - describing the dataset and making some initial plots

1) Describe the dataset, such that someone who happened upon this dataset could understand the variables and what they contain

The researchers are investigating whether pitch varies according to the setting a person is in - formal or informal. The experiment is performed on Koreans who now live in America. 
We have a dataset with 224 observations and seven coloums to describe each observation. The seven variable are: 
- Subject. All the matching "names" are one subject.
- Gender. Pretty self-explanatory.
- Scenario. The subjects are presented to seven different scenarios.
- Attitude. Whether it is the formal or informal condition of the scenario. Each participant goes through both. 
- Total duration. Amount of time the subject uses to "complete" the scenario. 
- f0mn. Mean pitch (in Hz).
- hiss count. Some sound might say while speaking (loud breath intakes). 

    i. Also consider whether any of the variables in _politeness_ should be encoded as factors or have the factor encoding removed. Hint: ```?factor```  
Gender and attitude. 
```{r}
ls.str(politeness)

politeness$gender <- as.factor(politeness$gender)
politeness$attitude <- as.factor(politeness$attitude)
politeness$scenario <- as.factor(politeness$scenario)
```


2) Create a new data frame that just contains the subject _F1_ and run two linear models; one that expresses _f0mn_ as dependent on _scenario_ as an integer; and one that expresses _f0mn_ as dependent on _scenario_ encoded as a factor
```{r}
F1df1 <- politeness %>% 
  filter(subject == "F1")
F1df1$scenario <- as.integer(F1df1$scenario)

F1df2 <- politeness %>% 
  filter(subject == "F1")
F1df2$scenario <- as.factor(F1df2$scenario)

F1_integer <- lm(f0mn ~ scenario, F1df1)
F1_factor <- lm(f0mn ~ scenario, F1df2)
summary(F1_integer)
summary(F1_factor)

```

    i. Include the model matrices, $X$ from the General Linear Model, for these two models in your report and describe the different interpretations of _scenario_ that these entail
```{r}
model.matrix(F1_integer)
model.matrix(F1_factor)

```
Interpretation when scenario is integer:
Scenario is treated as something continuous that can have any value on a number scale. 

Interpretation when scenario is factor:
Scenario is treated as something where we have a limited set of possible values (specifically the seven scenarios we see in the data set). So a scenario either is one of the seven or not, and R gives us a TRUE/FALSE (where 1 is true and 0 is false) answer to all of the seven possibilities.


    ii. Which coding of _scenario_, as a factor or not, is more fitting?
As factor. 

3) Make a plot that includes a subplot for each subject that has _scenario_ on the x-axis and _f0mn_ on the y-axis and where points are colour coded according to _attitude_
```{r}
plot_allparticipants <- politeness %>% 
  ggplot(aes(x = scenario, y = f0mn, color = attitude))+
  geom_point()+
  facet_wrap(~subject)
plot_allparticipants
```

    i. Describe the differences between subjects
Well, first of all the males have a way lower f0mn, which is very much expected. Other than that, a few of the participants vary way more in f0mn according to the scenario than others. It looks like the informal pitch is generally a liiiiittle bit higher than the formal one. 
    
## Exercise 2  - comparison of models

For this part, make sure to have `lme4` installed.  
You can install it using `install.packages("lme4")` and load it using `library(lme4)`  
`lmer` is used for multilevel modelling
```{r}
pacman::p_load("lme4")
```


```{r}
#mixed.model <- lmer(formula=..., data=...)
#example.formula <- formula(dep.variable ~ first.level.variable + (1 | second.level.variable))

```

1) Build four models and do some comparisons
    i. a single level model that models _f0mn_ as dependent on _gender_
    ii. a two-level model that adds a second level on top of i. where unique intercepts are modelled for each _scenario_
    iii. a two-level model that only has _subject_ as an intercept 
    iv. a two-level model that models intercepts for both _scenario_ and _subject_
```{r}
pacman::p_load(lmerTest)
model1 <- lm(f0mn~gender, politeness)
model2 <- lmerTest::lmer(f0mn~gender + (1|scenario), politeness)
model3 <- lmerTest::lmer(f0mn~gender + (1|subject), politeness)
model4 <- lmerTest::lmer(f0mn~gender + (1|scenario) + (1|subject), politeness)
```

    v. which of the models has the lowest residual standard deviation, also compare the Akaike Information Criterion `AIC`?
```{r}
summary(model1) #Residual standard error: 39.46 on 210 degrees of freedom. The sigma-function shows the same thing... 
sigma(model1)
summary(model2) #Residual standard deviation: 38.448
summary(model3) #Residual standard deviation: 32.04
summary(model4) #Residual standard deviation: 30.658

AIC(model1, model2, model3, model4) #2163.971, 2151.314, 2099.626, 2092.482

```
Both the residual standard deviation and the AIC get better for each model. So number 4 is the best one, judging from this. 

    vi. which of the second-level effects explains the most variance?
The jump from model1 to model3 is bigger than the jump from model1 to model2, so that would be subject. That also makes sense so me, that the difference in pitch is greater between people than between scenarios. But the best model includes both at the same time to explain the most. 

2) Why is our single-level model bad?
Because it doesn't explain a lot of variance. Just by looking at the raw data from task 3, we can see that it makes sense to incorporate random effects. E.g. not all women generally have the same pitch, just because they are the same gender. 

    i. create a new data frame that has three variables, _subject_, _gender_ and _f0mn_, where _f0mn_ is the average of all responses of each subject, i.e. averaging across _attitude_ and_scenario_
```{r}
politeness_meanpitch <- politeness %>% 
  select("subject", "gender", "f0mn")

politeness_meanpitch <- politeness_meanpitch %>% 
  group_by(subject, gender) %>% 
  summarize(mean(f0mn))

politeness_meanpitch$`mean(f0mn)`
```

    ii. build a single-level model that models _f0mn_ as dependent on _gender_ using this new dataset
```{r}
model_meanpitch <- lm(`mean(f0mn)`~gender, politeness_meanpitch)
summary(model_meanpitch)
```
  
    iii. make Quantile-Quantile plots, comparing theoretical quantiles to the sample quantiles) using `qqnorm` and `qqline` for the new single-level model and compare it to the old single-level model (from 1).i). Which model's residuals ($\epsilon$) fulfil the assumptions of the General Linear Model better?)
```{r}
pacman::p_load(car)

qqPlot(model_meanpitch)
qqPlot(model1)
```
The new model? I see less of a pattern, and the datapoints are within the dotted lines (which I don't really know what are).

    iv. Also make a quantile-quantile plot for the residuals of the  multilevel model with two intercepts. Does it look alright?
```{r}
qqPlot(residuals(model4))
```
Not really sure... To the right it looks a bit weird. 

3) Plotting the two-intercepts model
    i. Create a plot for each subject, (similar to part 3 in Exercise 1), this time also indicating the fitted value for each of the subjects for each for the scenarios (hint use `fixef` to get the "grand effects" for each gender and `ranef` to get the subject- and scenario-specific effects)
```{r}
grand_effects <- fixef(model4)
specific_effects <- ranef(model4)
grand_effects
specific_effects

ggplot(data = politeness, aes(x = scenario, y = f0mn, color = attitude)) +
  geom_point() +
  geom_point(aes(x = scenario, y = fitted(model4)), color = "black", shape = 18)+
  facet_wrap(~subject)

```
I'm a bit confused, since we don't include attitude in any of the models, so the fitted values will not take this into acconut.  
    
## Exercise 3 - now with attitude

1) Carry on with the model with the two unique intercepts fitted (_scenario_ and _subject_).
    i. now build a model that has _attitude_ as a main effect besides _gender_
```{r}
model5 <- lmerTest::lmer(f0mn~gender + attitude + (1|scenario) + (1|subject), politeness)
summary(model5)
```

    ii. make a separate model that besides the main effects of _attitude_ and _gender_ also include their interaction
```{r}
model6 <- lmerTest::lmer(f0mn~gender*attitude + (1|scenario) + (1|subject), politeness)
summary(model6)
```
  
    iii. describe what the interaction term in the model says about Korean men's pitch when they are polite relative to Korean women's pitch when they are polite (you don't have to judge whether it is interesting)
Don't quite know how to interpret the interaction... But it seems that men lower their pitch less in formal situations than women? 

2) Compare the three models (1. gender as a main effect; 2. gender and attitude as main effects; 3. gender and attitude as main effects and the interaction between them. For all three models model unique intercepts for _subject_ and _scenario_) using residual variance, residual standard deviation and AIC. 
```{r}
#Residual variance
summary(model4) #939.91
summary(model5) #882.7
summary(model6) #885.5

#Residual standard deviation
sigma(model4) #30.658
sigma(model5) #29.711
sigma(model6) #29.757

#AIC
AIC(model4, model5, model6) #2092.5, 2077.1, 2072.6

anova(model4, model5, model6) #well, model5 is significantly better than model4, I guess? 
```
Seems like model 5 and 6 are very similar, so I would always penalize the more complicated one. Model4 is not far off though... But model5 includes attitude, which is pretty relevant to the research questions, sooooo that's gonna be hard to come around. 

3)  Choose the model that you think describe the data the best - and write a short report on the main findings based on this model. At least include the following:
I choose model 5, since it's the simplest model that does the best in the comparison. Also, when making the models with lmerTest and then running summary, both of the main effects are significant, but the interaction isn't. 
  i. describe what the dataset consists of  
The model looks like this: model5 <- lmer(f0mn~gender + attitude + (1|scenario) + (1|subject), politeness)
It consists of two fixed effects being gender and attitude. This is where we expect systematic variance. We also include random intercepts based on scenario and subject. This is where we expect variance, that we can't model systematically. The model is made from the politeness dataset. 

  ii. what can you conclude about the effect of gender and attitude on pitch (if anything)?  
Not sure how to interpret IF we can conclude anything. Especially not about other people than the specific participants here... But if we can, then:
- Men generally have a pitch that is 115.4 Hz (with a standard error of 12.9) lower than womens.
- Being in the formal condition will lower pitch with 14.8 (standard error of 4.1) Hz. So the participants speak a bit deeper, when they are being polite. 

  iii. motivate why you would include separate intercepts for subjects and scenarios (if you think they should be included)  
I think both should be included, especially for subjects. I expect a lot of variance coming from the fact that people have way different baseline pitches. I also expect people to have different pitches based on condition - e.g. I would maybe speak lower/deeper in a library than in a classroom. But I expect this effect to have less impact than subject. 

  iv. describe the variance components of the second level (if any)  
The variance is the squared deviation. The exact numbers can be how to interpret, because they depend on the dataset, but we can see that a lot of the varieace is explained by the fixed and random effects, which is a good thing. 

  v. include a Quantile-Quantile plot of your chosen model
```{r}
qqnorm(residuals(model5))
```

